---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
```

Unziping the file. 

```{r}
if (!dir.exists("data")){
  dir.create("data")
}

if (!file.exists("activity.csv")){
  unzip("activity.zip", exdir = getwd())
}
```

Reading the data

```{r}
data0<- read.csv("activity.csv", header = TRUE)
```

Transforming the "date" column into Date class.

```{r}
data0$date <- ymd(data0$date)
```

Let's take a look at the data.

```{r}
head(data0)
```

## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day

```{r}
total_steps_day <- data0 %>% group_by(date) %>% 
  summarize(steps_day = sum(steps, na.rm = TRUE))

head(total_steps_day)
```

Make a histogram of the total number of steps taken each day

```{r}
hist(total_steps_day$steps_day, 
     main = "Total number of steps taken per day", 
     xlab = "Total steps taken per day", 
     col = "magenta", 
     ylim = c(0,20), 
     breaks = seq(0,25000, by=2500))
```

Calculate and report the mean of the total number of steps taken per day

```{r}
mean(total_steps_day$steps_day)
```

Calculate and report the median of the total number of steps taken per day

```{r}
median(total_steps_day$steps_day)
```

## What is the average daily activity pattern?

Calculating the averaged daily steps

```{r}
average_daily_steps <- aggregate(data0$steps, 
                                 by=list(data0$interval), 
                                 mean, 
                                 na.rm=TRUE)

names(average_daily_steps) <- c("interval", "average.steps")

head(average_daily_steps)
```

Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.

```{r}
plot(average_daily_steps$interval, average_daily_steps$average.steps, 
     type="l", 
     main = "Time series - average steps by intervals", 
     xlab= "Intervals", 
     ylab="Average steps", 
     lwd=2, 
     col="magenta")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
average_daily_steps[which.max(average_daily_steps$average.steps), ]$interval
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs.

```{r}
sum(is.na(data0$steps))
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Let's create a function that will retain the mean value for each interval.

```{r}
get.mean.value <- function(interval){
  average_daily_steps[average_daily_steps$interval == interval, ]$average.steps
}
```

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
data1 <- data0

for (i in 1:nrow(data1)){
    if(is.na(data1[i,]$steps)){
        data1[i,]$steps <- get.mean.value(data1[i,]$interval)
    }
}

head(data1)

sum(is.na(data1$steps))
```

Calculate the total number of steps taken per day in the new data with no missing values.

```{r}
total_steps_day_data1 <- data1 %>% group_by(date) %>% 
  summarize(steps_day = sum(steps, na.rm = TRUE))

head(total_steps_day_data1)
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r}
hist(total_steps_day_data1$steps_day, 
     main = "Total number of steps taken per day", 
     xlab = "Total steps taken per day", 
     col = "magenta", 
     ylim = c(0,20), 
     breaks = seq(0,25000, by=2500))
```

Calculate and report the mean of the total number of steps taken per day

```{r}
mean(total_steps_day_data1$steps_day)
```

Calculate and report the median of the total number of steps taken per day

```{r}
median(total_steps_day_data1$steps_day)
```

Do these values differ from the estimates from the first part of the assignment? 
Let's take a look at the two different means. 

```{r}
mean1 <- mean(total_steps_day$steps_day) #with NA values
mean2 <- mean(total_steps_day_data1$steps_day) #without NA values
percentage.diff.mean <- (mean2*100/mean1)-100
percentage.diff.mean
```

Let's take a look at the two different medians. 

```{r}
median1 <- median(total_steps_day$steps_day) #with NA values
median2 <- median(total_steps_day_data1$steps_day) #without NA values
percentage.diff.median <- (median2*100/median1)-100
percentage.diff.median
```

What is the impact of imputing missing data on the estimates of the total daily number of steps?

With inputed values, the mean went up by 15% and the median went up by 3.5%

## Are there differences in activity patterns between weekdays and weekends?

Let's include a new column (weekday) in the dataset.

```{r}
weekday <- weekdays(data1$date)
data1 <- cbind(data1,weekday)
head(data1)
```

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
data1$datetype <- sapply(data1$date, function(x) {
        if (weekdays(x) == "Saturday" | weekdays(x) =="Sunday") 
                {y <- "Weekend"} else 
                {y <- "Weekday"}
                y
        })

head(data1)
str(data1)
```

Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.

```{r}
steps_by_date <- aggregate(steps~interval + datetype, data1, mean, na.rm = TRUE)

plot<- ggplot(steps_by_date, aes(x = interval , y = steps, color = datetype)) +
       geom_line() +
       labs(title = "Average daily steps by type of date", x = "Interval", y = "Average number of steps") +
       facet_wrap(~datetype, ncol = 1, nrow=2)
print(plot)
```


